---
title: "PA1_template"
author: "Pantelia"
date: "2026-05-04"
output: html_document
---
##1. Code for reading in the dataset and/or processing the data
#Load the libraries and save the data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(lattice)

data <- read.csv("activity.csv")
data$date <- as.Date(data$date)

```
#Create a figure folder
```{r setup2, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "figure/"
)
```
##2. Histogram of the total number of steps taken each day
```{r, echo=TRUE}
daily_steps <- aggregate(steps ~ date, data, sum, na.rm = TRUE)
```
#Add Histogram
```{r, echo=TRUE}
hist(daily_steps$steps,
     main = "Total Steps per Day",
     xlab = "Steps")
```
##3.Calculate mean and Median number of steps taken each day
```{r, echo=TRUE}
mean(daily_steps$steps)
median(daily_steps$steps)
```
##4. Time series plot of the average number of steps taken 
-Average steps per interval
```{r, echo=TRUE}
interval_avg <- aggregate(steps ~ interval, data, mean, na.rm = TRUE)
```


#Plot Time series
```{r, echo=TRUE}
plot(interval_avg$interval,
     interval_avg$steps,
     type = "l",
     xlab = "Interval",
     ylab = "Average Steps",
     main = "Average Daily Activity Pattern")
```   

##5. The 5-minute interval that, on average, contains the maximum number of steps. 
-Max interval and count missing values
```{r, echo=TRUE}
interval_avg[which.max(interval_avg$steps), ]
sum(is.na(data$steps))
```
##6. Code to describe and show a strategy for imputing missing data.
-Imputation strategy: mean per interval
```{r imputation, echo=TRUE}
interval_means <- aggregate(steps ~ interval, data, mean, na.rm = TRUE)

data_imputed <- data

for(i in 1:nrow(data_imputed)) {
  if(is.na(data_imputed$steps[i])) {
    data_imputed$steps[i] <- interval_means$steps[
      interval_means$interval == data_imputed$interval[i]
    ]
  }
}
```
##7. Histogram of the total number of steps taken each day after missing values are imputed. Show mean and medians too
```{r, echo=TRUE}
daily_imputed <- aggregate(steps ~ date, data_imputed, sum)
hist(daily_imputed$steps,
     main = "Total Steps per Day (Imputed)",
     xlab = "Steps")
mean(daily_imputed$steps)
median(daily_imputed$steps)
```

#As you can see mean and median slightly changed after imputation 
##8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
- first create a factor variable foe weekday and weekend
```{r, echo=TRUE}
data_imputed$date <- as.Date(data_imputed$date)

data_imputed$daytype <- ifelse(
  weekdays(data_imputed$date) %in% c("Saturday", "Sunday"),
  "weekend",
  "weekday"
) 
```
#Secondly, create average number of steps taken per 5-minute interval across type

```{r avg_pattern, echo=TRUE}
avg_pattern <- aggregate(
  steps ~ interval + daytype,
  data = data_imputed,
  FUN = mean
)
```
# Finally create the panel plot to compare the average number of steps taken per 5-minute interval accross type

```{r, echo=TRUE}
library(lattice)

xyplot(
  steps ~ interval | daytype,
  data = avg_pattern,
  type = "l",
  layout = c(1, 2),
  xlab = "Interval",
  ylab = "Average Steps"
)
```